Sequence-guided pangenome graph layouts

### 2023-06-30 manuscript/main.tex
remove FIXME
titles
add citation and tiny tweaks

    \title[Pangenome graph layout by HOGWILD! Path-Guided Stochastic Gradient Descent]{Pangenome graph layout by HOGWILD! Path-Guided Stochastic Gradient Descent}
    \author[Heumos, Guarracino \textit{et~al}.]{
        Simon~Heumos\,$^{\orcidlink{0000-0003-3326-817X}1,2,\dagger}$,
        Andrea~Guarracino\,$^{\orcidlink{0000-0001-9744-131X}3,4,\dagger}$,
        Jan-Niklas Manuel Schmelzle\,$^{\orcidlink{0000-0001-8566-4049}5,6}$,
        Jiajie Li\,$^{\orcidlink{0000-0002-9467-0659}6}$,
        Zhiru Zhang\,$^{\orcidlink{0000-0002-0778-0308}6}$,
        Sven Nahnsen\,$^{\orcidlink{0000-0002-4375-0691}1,2}$,
        Pjotr Prins\,$^{\orcidlink{0000-0002-8021-9162}3}$,
        Erik~Garrison\,$^{\orcidlink{0000-0003-3821-631X}\text{\sfb 3},*}$
    }

@@

	\abstract{
	\textbf{Motivation:}
	The increasing availability of high quality genomes demands models to study genomic variability within entire populations.
	Pangenome graphs can represent the full genetic diversity between multiple genomes, but they may exhibit complex structures due to common, nonlinear patterns of genome variation and evolution.
	These structures introduce difficulty in downstream analyses, visualization, and interpretation. \\
	\textbf{Results:}
	To address these challenges, we propose a novel graph layout algorithm, the Path-Guided Stochastic Gradient Descent (PG-SGD).
	Our HOWGWILD! method uses the genomes, represented in the pangenome graph as paths, to move pairs of nodes in parallel.
	We show that our implementation efficiently computes the layout of gigabase-scale pangenome graphs, revealing their biological features. \\
	\textbf{Availability:}
	The PG-SGD algorithm has been integrated in \textit{ODGI} which is released as free software under the MIT open source license.
	Source code is available at \url{https://github.com/pangenome/odgi} and documentation at \url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}. \\
	\textbf{Contact:} \href{egarris5@uthsc.edu}{egarris5@uthsc.edu} \\
	\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
	}

@@

	Our implementation is multithreaded, threads can perform term update calculations in parallel. Layout values are stored in an atomic integer vector presenting the only multithreading bottleneck. Effectively, this presents a modified version of the HOGWILD! SGD algorithm which just writes the layout updates to a shared non-atomic integer vector. We went for an atomic vector, because this ensures memory saftey and in our experience there was basically no performance loss. Algorithm~\ref{alg:pg_sgd}.

	Due to the fact that we have a very high learning rate, our 1D implementation can work with a random layout as input. Though, by default, it just makes the given node order the starting layout. In 2D, we offer several random layout initializations: 
		\begin{enumerate*}[label=(\roman*)]
		\item Uniform,
		\item gaussian,
		\item or hilbert.
	\end{enumerate*}
	\FIXME{1D: random, node order in the input graph; 2D: Random (uniform or gaussian), hilbert}

@@

    \subsection{Performance}
	\FIXME{Describe the dataset (HPRC chromosomes), runtime and memory usage (specifying the faster/slower chromosome pangenome graph layout and the thinner/fatter in memory), just text no with a table}
    \FIXME{A comparison with Bandage might be worth, to give readers an idea of the scalability}

    \subsection{Pangenome graphs layouts reveal biology features}
	\FIXME{With the HPRC dataset, Describe a full chromosome, and 1 or a few selected loci, HLA, C4, beta-defensin locus, LPA, stuff from the HPRC paper or PGR-TK.
	In the supp a huge picture with a "karyotype might be hot}

@@

	The PG-SGD algorithm can be extended to any number of dimensions.
	Projecting a graph in a low number of dimensions, but still preserving its biologically relevant information, enable the application of algorithms that use the graph layout for variant detection and classification. 
	Helpful could be to integrate more layout objectives~\FIXME{Add citation for \url{https://arxiv.org/pdf/2112.01571.pdf}} like neighborhood preservation.

@@

	\begin{appendices}
	    \section{Appendix}
	    \FIXME{@ANDREA: Please add some more details about your clever quantization. Why. How. Where. When. Secs.} \\
	    \FIXME{PGR-TK Paper - Layouts of challenging regions of the human pangenome. - https://www.nature.com/articles/s41592-023-01914-y} \\
	    \FIXME{From 1D layout to 1D node order} \\
	    \FIXME{reference-guided path-guided stochastic gradient descent}
	\end{appendices}

>>>>

    %\title[Pangenome graph layout by HOGWILD! Path-Guided Stochastic Gradient Descent]{Pangenome graph layout by HOGWILD! Path-Guided Stochastic Gradient Descent}
    %\title[Pangenome graph layout by Path-Guided Stochastic Gradient Descent]{Pangenome graph layout by Path-Guided Stochastic Gradient Descent}
    %\title[Unveiling genome variation by sequence-guided pangenome graph layouts]{Unveiling genome variation by sequence-guided pangenome graph layouts}
    %\title[PG-SGD: sequence-guided pangenome graph layouts]{PG-SGD: sequence-guided pangenome graph layouts}
    %\title[Genome-guided pangenome graph layouts via Stochastic Gradient Descent]{Genome-guided pangenome graph layouts via Stochastic Gradient Descent}
	%\title[Sequence-guided pangenome graph layouts via Stochastic Gradient Descent]{Sequence-guided pangenome graph layouts via Stochastic Gradient Descent}
    %\title[Genome-guided pangenome graph layouts]{Genome-guided pangenome graph layouts}
	\title[Sequence-guided pangenome graph layouts]{Sequence-guided pangenome graph layouts}

	\author[Heumos, Guarracino \textit{et~al}.]{
        Simon~Heumos\,$^{\orcidlink{0000-0003-3326-817X}1,2,\dagger}$,
        Andrea~Guarracino\,$^{\orcidlink{0000-0001-9744-131X}3,4,\dagger}$,
        Jan-Niklas Manuel Schmelzle\,$^{\orcidlink{0000-0001-8566-4049}5,6}$,
        Jiajie Li\,$^{\orcidlink{0000-0002-9467-0659}6}$,
        Zhiru Zhang\,$^{\orcidlink{0000-0002-0778-0308}6}$,
        Sven Nahnsen\,$^{\orcidlink{0000-0002-4375-0691}1,2}$,
        Pjotr Prins\,$^{\orcidlink{0000-0002-8021-9162}3}$,
        Erik~Garrison\,$^{\orcidlink{0000-0003-3821-631X}\text{\sfb 3},*}$
    }

@@

	\abstract{
	\textbf{Motivation:}
	The increasing availability of complete genomes demands models to study genomic variability within entire populations.
	Pangenome graphs can represent the full genetic diversity between multiple genomes, but they may exhibit complex structures due to common, nonlinear patterns of genome variation and evolution.
	These structures introduce difficulty in downstream analyses, visualization, and interpretation. \\
	\textbf{Results:}
	To address these challenges, we propose a novel graph layout algorithm, the Path-Guided Stochastic Gradient Descent (PG-SGD).
	Our HOWGWILD! method uses the genomes, represented in the pangenome graph as paths, to move pairs of nodes in parallel.
	We show that our implementation efficiently computes the layout of gigabase-scale pangenome graphs, revealing their biological features. \\
	\textbf{Availability:}
	The PG-SGD algorithm has been integrated in \textit{ODGI} which is released as free software under the MIT open source license.
	Source code is available at \url{https://github.com/pangenome/odgi} and documentation at \url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}. \\
	\textbf{Contact:} \href{egarris5@uthsc.edu}{egarris5@uthsc.edu} \\
	\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
	}

@@

	Our implementation is multithreaded, threads can perform term update calculations in parallel. Layout values are stored in an atomic integer vector presenting the only multithreading bottleneck. Effectively, this presents a modified version of the HOGWILD! SGD algorithm which just writes the layout updates to a shared non-atomic integer vector. We went for an atomic vector, because this ensures memory saftey and in our experience there was basically no performance loss.

	Due to the fact that we have a very high learning rate, our 1D implementation can work with a random layout as input. Though, by default, it just makes the given node order the starting layout. In 2D, we offer several random layout initializations: 
		\begin{enumerate*}[label=(\roman*)]
		\item Uniform,
		\item gaussian,
		\item or Hilbert.
	\end{enumerate*}

@@

    \subsection{Performance}
	\FIXME{Describe the dataset (HPRC chromosomes), runtime and memory usage (specifying the fastest/slowest chromosome pangenome graph layout and the thinnest/fattest in memory), just text no with a table}
    \FIXME{A comparison with Bandage might be worth, to give readers an idea of the scalability}

    \subsection{Pangenome graphs layouts reveal biology features}
	\FIXME{With the HPRC dataset, Describe a full chromosome, and 1 or a few selected loci, HLA, C4, beta-defensin locus, LPA, stuff from the HPRC paper or PGR-TK.
	In the supp a huge picture with a "karyotype", that is all chromosome layout together sorted by chromosome number, might be hot}

@@

	The PG-SGD algorithm can be extended to any number of dimensions.
	Projecting a graph in a low number of dimensions, but still preserving its biologically relevant information, enable the application of algorithms that use the graph layout for variant detection and classification. 
	Helpful could be to integrate more layout objectives \citep{ahmed_multicriteria_2021} like neighborhood preservation.

@@

	\begin{appendices}
	    \section{Appendix}
	    %\FIXME{@ANDREA: Please add some more details about your clever quantization. Why. How. Where. When. Secs.} \\
	    %\FIXME{PGR-TK Paper - Layouts of challenging regions of the human pangenome. - https://www.nature.com/articles/s41592-023-01914-y} \\
	    \FIXME{From 1D layout to 1D node order} \\
	    \FIXME{reference-guided path-guided stochastic gradient descent}
	\end{appendices}
